Deterministic Bayesian inference methods for the Naomi model

### 2023-03-28 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
Iterate slides #24

* If you want to integrate something deterministically, you could use numerical integration, otherwise called \textcolor{hilit}{quadrature}
* Select nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ then compute the sum
$$
\tilde p(y) = \sum_{\vartheta \in \mathcal{Q}} p(y, \vartheta) \omega(\vartheta)
$$

@@

* Remember how integration is taught? (Rienmann sums)
* Try computing $\int_0^\pi \sin(x) \text{d}x = 2$ using trapezoid rule

```{r}
trapezoid_rule <- function(x, spacing) {
  w <- rep(spacing, length(x))
  w[1] <- w[1] / 2
  w[length(x)] <- w[length(x)] / 2
  sum(w * x)
}
```

@@

```{r message=FALSE, echo=FALSE}
library(tidyverse)
cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

@@

  plot + 
    geom_bar(
      data = data.frame(x = grid, y = sin(grid)),
      aes(x = x, y = y), alpha = 0.7, stat = "identity",
      inherit.aes = FALSE, fill = cbpalette[2]) +
    theme_minimal() +
    labs(
      subtitle = paste0("Number of quadrature points: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"),
      x = "x", y = "sin(x)"
    )
}

@@

```{r message=FALSE, echo=FALSE}
trapezoid_plot(N = 30)
```

@@

```{r message=FALSE, echo=FALSE}
trapezoid_plot(N = 100)
```

@@

* Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$
* If we set $\omega(\vartheta_i) = 1/N$ for all $i$ then we get a **Monte Carlo** (MC) estimate
$$
\tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i)
$$
* For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain which if you squint a bit is good enough (MCMC)

@@

@@

* Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized
  * If you're using $k$ points per dimension it's $k^{\text{dim}(\vartheta)}$
* Previously all of the parameters were under the symbol $\vartheta$ -- what if we split them up as being $\vartheta = (x, \theta)$
* The key part about this is that $\dim(x) = N$ is big and $\dim(\theta) = m$ is small

@@

* There is nothing inherently special about spatio-temporal statistics, but we do often end up in similar situations because of the structures of the problems we tackle
* We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$
* Usually we associate parameters to spatio-temporal locations as well as observations
* This ends up with us having something like $\{x_{s, t}\}$

@@

1. There might be \textcolor{hilit}{a lot} of spatio-temporal locations, so $\text{dim}(x)$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
2. Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]
* Not IID anymore!

@@

* Hey wait a second, is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian?
* No, not in general. But...

1. We just described a class of models (LGMs) where some subset of the parameters ($x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian
2. We just talked about how big $x$ is in comparison to $\theta$! $\implies$ most of the work in our integral can be done using a **marginal Laplace** approximation to get rid of $x$

@@

* Now we can compute $p_{\texttt{LA}}(\theta, y)$ but what we really want is still $p(y)$
* But hopefully^[Really: hopefully] the dimension of $\theta$ is small enough that we can now tackle this with quadrature
* So pick some nodes $\mathcal{Q}$ and a weighting function $\omega$ and away we go
$$
p(y) \approx \sum_{\theta \in \mathcal{Q}} p_{\texttt{LA}}(\theta, y) \omega(\theta)
$$
* This is the famous integrated nested Laplace approximation (INLA)

@@

* `TMB` [@kristensen2015tmb] is an R package which implements the Laplace approximation for latent variable models
* To use `TMB`, you write your objective function $- \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)$ in `TMB`'s C++ syntax
* For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks as follows

@@

* It runs quickly, and is flexible enough to write the model
* Another answer: we don't have any better options
  * Markov chain Monte Carlo (MCMC) is accurate (eventually) but takes too long
  * Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a latent Gaussian model, and isn't compatible with the `R-INLA` implementation of INLA

@@

* Implement an algorithm similar to INLA which
1. is compatible with Naomi
2. uses `TMB` to perform Laplace approximation 
* Does this improves the quality of the inferences we get?

@@

* Alex Stringer in Toronto $\to$ Waterloo had been thinking along similar lines, and made a lot of progress

1. Implementing an algorithm similar to INLA, using a specific quadrature rule $\mathcal{Q}$ called adaptive Gauss-Hermite quadrature (AGHQ) which he and others argue should be the default for this problem
2. Defining a class of models called extended latent Gaussian models (ELGMs) which Naomi fits into

@@

* The nodes and weights we use should probably depend on the integrand
  * Especially when the integrand is also a function of $y$ as in $p(y, \vartheta)$
* Let $z \in \mathcal{Q}(m, k)$ then $\theta(z) = \hat \theta + Lz$ where $L$ is the lower Cholesky of $H(\hat \theta)$

@@

```{r, echo=FALSE, message=FALSE, fig.cap="Rotate by the lower Cholesky."}
gg3 <- gg
mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)

>>>>

* If you want to integrate something deterministically, you could use numerical integration, otherwise called \textcolor{hilit}{quadrature}
* Choose nodes $\vartheta \in \mathcal{Q} \subset \varTheta$ and weights $\omega: \varTheta \to \mathbb{R}$ and compute the weighted sum
$$
\tilde p(y) = \sum_{\vartheta \in \mathcal{Q}} p(y, \vartheta) \omega(\vartheta)
$$

@@

* Remember how integration is taught? (Rienmann sums)
* Try computing $\int_0^\pi \sin(x) \text{d}x = 2$ using trapezoid rule
* Nodes evenly spaced through $[0, \pi]$

```{r}
trapezoid_rule <- function(x, spacing) {
  w <- rep(spacing, length(x)) # Weights given by space between nodes
  w[1] <- w[1] / 2 # Apart from the first which is halved
  w[length(x)] <- w[length(x)] / 2 # And the last, also halved
  sum(w * x) # Compute the weighted sum
}
```

@@

```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."}
library(tidyverse)
cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

@@

  plot + 
    geom_bar(
      data = data.frame(x = grid, y = sin(grid)),
      aes(x = x, y = y), alpha = 0.7, stat = "identity",
      inherit.aes = FALSE, fill = cbpalette[2]) +
    theme_minimal() +
    labs(
      subtitle = paste0("Number of nodes: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"),
      x = "x", y = "sin(x)"
    )
}

@@

```{r message=FALSE, echo=FALSE, fig.cap="With 30 nodes it's 0.002 off."}
trapezoid_plot(N = 30)
```

@@

```{r message=FALSE, echo=FALSE, fig.cap="With 100 nodes it's pretty much correct."}
trapezoid_plot(N = 100)
```

@@

* Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$
* The Monte Carlo (MC) estimate is
$$
\tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i)
$$
which is quadrature with nodes given by the samples and $\omega(\vartheta_i) = 1/N$ for all $i$
* For complicated models^[Or not even that complicated] it's not possible to sample directly from $p(y, \vartheta)$, but we can usually sample from a Markov chain (MCMC)

@@

# Curse of dimensionality

* Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized, because there is an exponential increase in the volume you need to cover
* If you're using $k$ points per dimension it's $k^{\text{dim}(\vartheta)}$ e.g. $k = 5$ then

```{r}
5^{1:8}
```

@@

* Previously all of the parameters were under the symbol $\vartheta$ -- what if we split them up as being $\vartheta = (x, \theta)$
* The key part about this is that $\dim(x) = N$ is big and $\dim(\theta) = m$ is small

@@

* There is nothing inherently special about spatio-temporal statistics, but we do often end up tackling problems with similar structures
* We have observations indexed by space $s \in \mathcal{S}$ and time $t \in \mathcal{T}$
* Usually we associate parameters to spatio-temporal locations, giving us something like $\{x_{s, t}\}_{s \in \mathcal{S}, t \in \mathcal{T}}$

@@

1. There might be \textcolor{hilit}{a lot} of spatio-temporal locations, so $N$ might be pretty big! If you have 100 districts and 10 years, that's already $100 \times 10 = 1000$ parameters
2. Perhaps we're willing to make quite strong assumptions about how things vary over space-time^[Are there any slides about spatial statistics that don't describe Tobler's first law of geography?]: not IID anymore!

@@

* Hey wait a second, is it reasonable to just assume $p(\vartheta \, | \, y)$ is Gaussian?
* There is the Bernstein–von Mises theorem, but we can't rely on that in general. However...

1. We just described a class of models (LGMs) where some subset of the parameters ($x$) have a Gaussian prior $\implies$ it's a lot more reasonable to think that they would have a marginal posterior which is close to Gaussian
2. We just talked about how big $x$ is in comparison to $\theta$! $\implies$ most of the work in our integral can be done using a \textcolor{hilit}{marginal Laplace} approximation to get rid of $x$

@@

* Now we can compute $p_{\texttt{LA}}(\theta, y)$ but what we really want is still $p(y)$
* But hopefully^[Really: hopefully] $m$ is small enough that we can now tackle this with quadrature
* So pick some nodes $\mathcal{Q}$ and a weighting function $\omega$ and away we go
$$
p(y) \approx \sum_{\theta \in \mathcal{Q}} p_{\texttt{LA}}(\theta, y) \omega(\theta)
$$
* This is the famous integrated nested Laplace approximation (INLA)

@@

* To use `TMB` [@kristensen2015tmb], you write your objective function $- \log p(y, x, \theta)$ in the C++ syntax
* For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks like (next slide)
* `TMB` implements the Laplace approximation! Set `random = "x"`

@@

* It runs quickly, and is flexible enough to write the model
* Another answer is that we don't have any better options

| Option | Would it work? |
|--------|--------------------|
| MCMC | No! It is accurate (eventually) but takes too long |
| INLA via `R-INLA`        | No! Though Naomi is a spatial model with a large Gaussian latent field, it isn't technically a LGM, and isn't compatible with `R-INLA` |

@@

* Implement an algorithm inspired by INLA which
1. is compatible with Naomi and its intricacies
2. uses `TMB` -- and thereby automatic differentiation -- to perform the Laplace approximation

@@

* Alex Stringer in Toronto $\to$ Waterloo had independently been thinking along similar lines, and made a lot of progress

1. Implementing an algorithm similar to INLA, using a specific quadrature rule $\mathcal{Q}$ called adaptive Gauss-Hermite quadrature (AGHQ) which arguably should be the default for this problem
2. Defining a class of models called extended latent Gaussian models (ELGMs) which Naomi fits into

@@

* The nodes and weights we use should probably depend on the integrand
* Especially when the integrand is also a function of $y$, which we don't know in advance, as in $p(y, \vartheta)$
  * i.e. how could any fixed quadrature rule be appropriate for all possible $y$?
* Let $z \in \mathcal{Q}(m, k)$ then
$$
\theta(z) = \hat \theta + Lz
$$
where $L$ is the lower Cholesky of $H = LL^\top$
* Other matrix decompositions can also be used e.g. spectral $H = E \Lambda E^\top = (E\Lambda^{1/2})(E\Lambda^{1/2})$
  * Arguably this is preferable: symmetric with respect to the principle axis [@jackel2005note]

@@

```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $L$."}
gg3 <- gg
mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)

```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2}$."}
gg3 <- gg
mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1)
